---
title: "Data Import"
output: github_document
---

This file is for doing data important

```{r}
library(tidyverse)
library(readxl)
library(haven)
```

#import 2 CSVs


Import our first dataset

```{r}
litters_df = 
  read_csv(
    file = "data/FAS_litters.csv", 
    na = c("", "NA", ".")
    )

#this dataset used both "NA" and "." to indicate missing data

litters_df = janitor::clean_names(litters_df)

#added unscores between words of variable names (snake case)
```
Import second dataset

```{r}
pups_df = 
  read_csv(
    "data/FAS_pups.csv", 
           skip=3, 
           na = c("", "NA", ".")
  )

# do not need to say file = because it assumes that
# skip = n skips n number of rows at the top when importing the data set

pups_df = janitor::clean_names(pups_df)
```
## Look at the data 

```{r}
litters_df
pups_df
```

Head or tail
```{r}
head(litters_df, 20)

tail(pups_df, 5)

#head shows the first n amount of rows of the data set
#tail shows the last n amount of rows of the data set
```

Skimming?

```{r}
skimr::skim(pups_df)
```

## Oh excel...

Jenny Bryan made 'readxl' to solve our problems.

```{r}
mlb_df = 
  read_excel("data/mlb11.xlsx")
```

Look at the data!

```{r}
mlb_df
```

Load some LOTR data

Import FOTR words

```{r}
fotr_df = 
  read_excel(
    "data/LotR_Words.xlsx",
             range = "B3:D6"
             )

fotr_df

# range imports the chosen cells
```

What about Two Towers

```{r}
towers_df = 
  read_excel(
    "data/LotR_Words.xlsx",
             range = "F3:H6"
             )

towers_df
```

## Import SAS

Read in the PULSE data set

```{r}
pulse_df = 
  read_sas("data/public_pulse_data.sas7bdat")

pulse_df = janitor::clean_names(pulse_df)

pulse_df
```

## What's so bad about 'read.csv()'?

Import pups with 'read.csv()'

```{r}
pups_df_base =
  read.csv("data/FAS_pups.csv")

pups_df_base

#outputs as many rows as possible, looks bad

```
Won't show: lazy about variable names: super slow for big data sets.
