---
title: "Data Manipulation"
output: github_document
---

This file is for doing data manipulation

```{r}
library(tidyverse)
```

Load the litters and pups data sets

```{r}
litters_df = 
  read.csv("data/FAS_litters.csv",
           na = c("", "NA", "."))

litters_df = janitor::clean_names(litters_df)

pups_df =
  read.csv("data/FAS_pups.csv",
           skip = 3,
           na = c("", "NA", "."))

pups_df = janitor::clean_names(pups_df)
```

## `select` some variables

```{r}
select(litters_df, group, litter_number)

select(litters_df, group:gd_of_birth)

select(litters_df, -group)

select(litters_df, starts_with("gd"), group)

select(litters_df, GROUP=group, everything())
```

To use select, the first thing in the brackets in the data frame, and the next things are rhe variables you want to keep

Using : means to select all the variables between the two listed variables

Using - before the variable name means to keep all variables except for that one

The = renames the variable

everything means to keep everything else


A couple of alternatives
```{r}
rename(litters_df, GROUP = group)

relocate(litters_df, litter_number, gd0_weight)
```

relocate moves the listed variables to the front


In the pups data, select litter number, sex, and PD ears

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## `filter` some rows

```{r}
filter(litters_df, gd_of_birth == 20)

filter(litters_df, pups_born_alive <= 5)

filter(litters_df, pups_born_alive !=6)

filter(litters_df, group %in% c("Con7", "Con8"))

filter(litters_df, group %in% c("Con7", "Con8"), pups_born_alive < 7)

drop_na(litters_df)

drop_na(litters_df, gd0_weight)

filter(litters_df, pups_born_alive < 5 | pups_born_alive >8)

# != is not equal to

# drop_na gets rid of all of the rows with missing values. You can specify only get rid of rows with missing values of a certain variable by adding , variable name after the data set name

# the | means or
```


In the pups dataset, keep only rows where PD walk is less than 11 and sex is 2
```{r}
filter(pups_df, pd_walk < 11, sex ==2)
```


## `mutate` to change some columns

```{r}
mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight,
  wt_gain_squared = wt_gain ^2,
  group = str_to_lower(group))

# this is creating a variable called wt_gain where it equals the value of gd18_weight minus the value of gd0_weight

# str_to_lower() makes the selected variable name all lowercase
```

In the pups dataset, create a variable that subtracts 7 from pd walk 

```{r}
mutate(pups_df,
       new_pd = pd_walk - 7)
```


## `arrange` the data

```{r}
arrange(litters_df, pups_born_alive)

# sorts it by ascending of the selected variable

arrange(litters_df, desc(pups_born_alive))

# sorts it by descending of the selected variable
```

This is bad:
```{r}
litters_df = read_csv("data/FAS_litters.csv", na = c("", "NA", "."))
litters_rename_df = janitor::clean_names(litters_df)
litters_with_gd_df = select(litters_rename_df, group, starts_with("gd"))
litters_no_na_df = drop_na(litters_with_gd_df)
litters_wt_gain_df = mutate(litters_no_na_df, wt_gain = gd18_weight - gd0_weight)

# don't do it like this because it is too confusing
```

This is worse:
```{r}
select(janitor::clean_names(read_csv("data/FAS_litters.csv", na = c("", "NA", "."))), group, starts_with("gd"))
```


This is good:
```{r}
litters_df =
  read_csv("data/FAS_litters.csv", na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  select(group, starts_with("gd")) |>
  drop_na() |>
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
  )

```


Load pups, clean names, drop missing, keep litter number and pd variables, add pd walk - 7

```{r}
pups_df =
  read_csv("data/FAS_pups.csv", 
           skip = 3,
           na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  drop_na() |>
  select(litter_number, starts_with("pd")) |>
  mutate(pd_walk_new = pd_walk - 7)
```

