---
title: "Data Manipulation"
output: github_document
---

This file is for doing data manipulation

```{r}
library(tidyverse)
```

Load the litters and pups data sets

```{r}
litters_df = 
  read.csv("data/FAS_litters.csv",
           na = c("", "NA", "."))

litters_df = janitor::clean_names(litters_df)

pups_df =
  read.csv("data/FAS_pups.csv",
           skip = 3,
           na = c("", "NA", "."))

pups_df = janitor::clean_names(pups_df)
```

## `select` some variables

```{r}
select(litters_df, group, litter_number)

select(litters_df, group:gd_of_birth)

select(litters_df, -group)

select(litters_df, starts_with("gd"), group)

select(litters_df, GROUP=group, everything())
```

To use select, the first thing in the brackets in the data frame, and the next things are rhe variables you want to keep

Using : means to select all the variables between the two listed variables

Using - before the variable name means to keep all variables except for that one

The = renames the variable

everything means to keep everything else


A couple of alternatives
```{r}
rename(litters_df, GROUP = group)

relocate(litters_df, litter_number, gd0_weight)
```

relocate moves the listed variables to the front


In the pups data, select litter number, sex, and PD ears

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

