---
title: "Tidy Data"
output: github_document
date: "2026-09-29"
---

This file is for doing data tidying

```{r}
library(tidyverse)
```

## Let's tidy some data!

```{r}
pulse_df = 
  haven::read_sas("data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names()

# haven loads datasets from sas
# we want bdi_score to be one columns, right now it is 4 different columns
```

# Okay, let's tidy
```{r, eval = FALSE}
pulse_tidy_df = 
  pulse_df |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi_score"
  ) |> 
  mutate(
    visit = replace(visit, visit=="bl", "00m") # bl = baseline
  )

# "bdi_score_bl:bdi_score_12m" means that all of the following code applies to all the variables in between and including the ones listed

# "names_to = "visit"" makes it so the names of those previous listed variables become the values of a new variable called visit

# "names_prefix = "bdi_score_"" removes that prefix (bdi_score_) from all the values in the new "visit" variable

# "values_to = "bdi_score"" makes it so the values of the previous listed variables are put into a new variable called bdi_score

# "mutate(visit = replace(visit, visit=="bl", "00m")" makes it so values of the variable visit that are equal to "bl" become equal to "00m"
```

Just use once...
```{r}
pulse_df = 
  haven::read_sas("data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi_score"
  ) |> 
  mutate(
    visit = replace(visit, visit=="bl", "00m") # bl = baseline
  )

# this does everything from the last chunk in one step instead of first loading the dataset and then tidying it
```

Let's practice!

In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18
```{r}
litters_df2 =
  read_csv( 
    file = "data/FAS_litters.csv", 
    na = c("", "NA", ".")
  ) |> 
  janitor::clean_names() |>
  select(litter_number, gd0_weight, gd18_weight) |> 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd",
    values_to = "weight"
  ) |> 
  mutate(
    gd = recode_values(
      gd,
      "gd0_weight" ~ 0,
      "gd18_weight" ~ 18
    )
  )
  
# this part "mutate(gd = recode_values(gd, "gd0_weight" ~ 0, "gd18_weight" ~ 18) makes it so, for the variable gd, the values that are gd0_weight should be renamed to 0 and the values that are gd18_weight should be renamed to 18
```

## Deliberately untidy data
```{r}
analysis_df = 
  tibble(
    groups = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean_outcome = c(4, 8, 3.5, 4.6)
  )
```

Let's untidy this dataset for human readability
```{r}
analysis_df |> 
  pivot_wider(
    names_from = time, 
    values_from = mean_outcome
  ) |> 
  knitr::kable()

# pivot_wider is the inverse of pivot_longer

# "names_from = time" makes it so the values of the variable "time" become new variables instead of the variable "time"

# "values_from = mean_outcome" makes it so the values from the variable "mean_outcome" become the values of the new variables "pre" and "post", which came from the old "time" variable

# "knitr::kable()" just makes the table look nicer in the output

```

### bind some rows

first, import each LOTR movie table

```{r}
fellowship_df = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "B3:D6") |> 
  mutate(movie = "fellowship")

two_towers_df = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "F3:H6") |> 
  mutate(movie = "two towers")

return_df = 
  readxl::read_excel("data/LotR_Words.xlsx", range = "J3:L6") |> 
  mutate(movie = "return of the king")
```

Next put all of these together and tidy

```{r}
lotr_df = 
  bind_rows(fellowship_df, two_towers_df, return_df) |> 
  janitor::clean_names() |> 
  relocate(movie) |> 
  pivot_longer(
    female:male,
    names_to = "gender",
    values_to = "words"
  )
```

## joion FAS datasets

import both

```{r}
pups_df = 
  read_csv("data/FAS_pups.csv",
           skip = 3,
           na = c(".", "", "NA")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      2 ~ "female"
    )
  )

litters_df = 
  read_csv("data/FAS_litters.csv",
           na = c(".", "", "NA")) |> 
  janitor::clean_names() |> 
  relocate(litter_number) |> 
  separate(group, into = c("dose", "day_of_tx"), 3) |>  #separates at position 3 (i.e. con and 7); could also do -1 or specify at _
  mutate(
    dose = str_to_lower(dose),
    day_of_tx = as.numeric(day_of_tx),
    gd_weight_gain = gd18_weight - gd0_weight
  )

fas_df = 
  left_join(pups_df, litters_df, by = "litter_number")
```

